---
title: "Pokemon"
author: "u2056802"
date: '2023-09-21'
output: pdf_document
---

This is an analysis for a dataset of all the pokemon from generations 1-7. 
The objective is to find what types are better choices according to different variables,
such as "number of weaknesses" as well as based on their stats like hp, attack, etc.

1. Dataset description

The dataset pokemon.csv is obtained from Kaggle (link)

It has 801 observations.

41 variables
  1. Abilities
  2-19 Type effectiveness
  20,26,29 Attack, Defense, HP
  34,35,36 Sp Attack, Sp Defense, Speed
  21. Base Egg steps
  22. Base happiness
  23. Base total
  24. Capture rate
  25. Classification
  27. Experience growth
  28. Height in meters
  30,31 Japanese name, (english) name
  32. Percentage male
  33. Pokedex number
  37,38 Type1, type2
  39. Weight in kg
  40. Generation
  41. Legendary (bool)


```{r}
library(ggplot2)
library(dplyr)

```


Data :

```{r}
rawdata <- read.csv("pokemon.csv")
cols <- c('yellow4','gray17','slateblue1','gold1','rosybrown1','tomato3','orangered1',
          'cadetblue1','slateblue4','green2','darkgoldenrod','turquoise','snow3',
          'purple','violetred1','tan4','gray79','royalblue1')
color_mapping <- c("flying" = "cadetblue1",
                   "fairy" = "rosybrown1",
                   "ice" = "turquoise",
                   "steel" = "gray79",
                   "dragon" = "slateblue1",
                   "ghost" = "slateblue4",
                   "fighting" = "tomato3",
                   "dark" = "gray17",
                   "ground" = "darkgoldenrod",
                   "poison" = "purple",
                   "electric" = "gold1",
                   "rock" = "tan4",
                   "fire" = "orangered1",
                   "psychic" = "violetred1",
                   "bug" = "yellow4",
                   "grass" = "green2",
                   "normal" = "snow3",
                   "water" = "royalblue1",
                   "none" = "beige")
```

Data cleaning

```{r}
data <- rawdata %>%
  #replace empty with none for type2
  mutate(type2 = ifelse(type2 == '' | type2==type1,"none",type2)) %>%
  #colors
  mutate(type1color = color_mapping[type1]) %>%
  mutate(type2color = color_mapping[type2]) 

```


EDA STAGE


```{r}
counts <- table(data$type1)
countdf <- as.data.frame(counts)
countdf <- countdf[order(countdf$Freq), ]
```

```{r}
barplot(countdf$Freq, horiz=TRUE,beside=TRUE,names.arg=countdf$Var1,col=color_mapping,
        cex.names=0.7, las=2, xlim=c(0,120),main="Frequency of Primary Type (gens 1-7)")

```

```{r}
counts2 <- table(data$type2)
countdf2 <- as.data.frame(counts2)
countdf2 <- countdf2[order(countdf2$Freq), ]
print(countdf2)
```

```{r}
cat(countdf2[which.max(countdf2$Freq), ]$Freq,"out of",nrow(data),"pokemon don't have a secondary type.", "\n")
cat(countdf2[countdf2$Var1 == "flying",]$Freq, "pokemon have flying as its secondary type.","\n")
```


ANALYSIS STAGE

type analysis 

Type and Weight correlation:

```{r}
total_mean <- mean(data$weight_kg, na.rm=TRUE)
mean_by_type <- aggregate(data$weight_kg, by = list(data$type1), FUN = mean, na.rm=TRUE)
mean_by_type[order(mean_by_type$x),]
```

Based on the rank statistics, the types dragon ground and steel have the highest average weight.

```{r}
summary(aov(data$weight_kg ~ data$type1 + data$type2))
```

```{r}
tukey <- TukeyHSD(aov(data$weight_kg ~ data$type1 + data$type2))
p_values_type1 <- tukey$`data$type1`[, "p adj"]


p_values_type2 <- tukey$`data$type2`[, "p adj"]


tukey_result_type1_sorted <- tukey$`data$type1`[order(p_values_type1), ]
low_p_values_type1 <- tukey_result_type1_sorted[tukey_result_type1_sorted[, "p adj"] <= 0.05, ]


tukey_result_type2_sorted <- tukey$`data$type2`[order(p_values_type2), ]
low_p_values_type2 <- tukey_result_type2_sorted[tukey_result_type2_sorted[, "p adj"] <= 0.05, ]


print("Low p-values for type1 (close to 0):")
print(low_p_values_type1)

print("Low p-values for type2 (close to 0):")
print(low_p_values_type2)

```

```{r}
tukey_row1 <- tukey$`data$type1`[, "p adj"]
tukey_row2 <- tukey$'data$type2'[, "p adj"]
tukey_row2


```



```{r}
data$type1 <- factor(data$type1, levels = unique(data$type1))
```

hp and defense by type

legendary analysis

- legendary influences weight, height, stats

```{r}
legendaries <- data[data$is_legendary == TRUE, ]
non_legendaries <- data[data$is_legendary == FALSE, ]

main_stats <- c("hp","attack","sp_attack","defense","sp_defense","speed")
non_main_stats <- c("height_m","weight_kg")

t_test <- lapply(c(main_stats,non_main_stats), function(var){
  test_res <- t.test(legendaries[[var]], non_legendaries[[var]])
  return(test_res)
})

names(t_test) <- c('hp', 'atk', 'sp.atk', 'def', 'sp.def', 'speed', 'height', 'weight')
t_test
```

Observing the differences of average for each of the stats, its evident that legendaries have over-buffed stats.



Number of weaknesses

```{r}
calculate_weakness_breakdown <- function(data) {
  # initialize variables
  unique_types <- character(0)
  weakness_counts <- integer(0)

  # iterate 
  for (i in 1:nrow(data)) {
    # type information
    type1 <- data$type1[i]
    type2 <- data$type2[i]

    #key for unique type or combination of types
    type_key <- paste(c(type1, type2), collapse = "-")

    # check if type has been encountered
    if (!(type_key %in% unique_types)) {
      # update set of encountered types
      unique_types <- c(unique_types, type_key)

      # identify columns "against_" and having a numeric value
      against_columns <- grep("^against_", names(data), value = TRUE)

      # sum where value higher or equal than 2
      weak_against_count <- sum(data[i, against_columns] >= 2, na.rm = TRUE)

      # append the results to the vectors
      weakness_counts <- c(weakness_counts, weak_against_count)
    }
  }

  # Create a data frame with the results
  result_df <- data.frame(Type = unique_types, WeaknessCount = weakness_counts)
  result_df <- result_df[order(result_df$WeaknessCount), ]
  
  return(result_df)
}
result <- calculate_weakness_breakdown(data)
print(result)
```

type effectiveness is one of the key points when playing pokemon. 
the average number of weaknesses a pokemon has is 3.766
The 7 strongest pokemon types, which have only one weakness, are:

  single types
electric (only weak to ground) x2
normal (only weak to fighting) x2
  dual types
normal-dark (only weak to fighting) x4
water-ground (only weak to grass) x4
bug-steel (only weak to fire) x4
dark-ghost (only weak to psychic) x4 Sableye
poison-dark (only weak to ground) x4 

The 6 weakest pokemon types, with 7 distinct type weaknesses, are:
rock-fighting (legendary Terrakion)
grass-ice 
rock-psychic
grass-dark
psychic-grass (legendary Celebi, Exeggcute and Exeggutor)
rock-dark (Tyranitar)

however the number of weaknesses is not always that important. 
For example, in the case of Tyranitar:

```{r}
hp_percentile <- ecdf(data$hp)
attack_percentile <- ecdf(data$attack)
sp_attack_percentile <- ecdf(data$sp_attack)
defense_percentile <- ecdf(data$defense)
sp_defense_percentile <- ecdf(data$sp_defense)
speed_percentile <- ecdf(data$speed)

print(paste("HP", data[248,"hp"], "is in the", round(hp_percentile(data[248,"hp"]) * 100, 2), "th percentile"))
print(paste("Attack", data[248,"attack"], "is in the", round(attack_percentile(data[248,"attack"]) * 100, 2), "th percentile"))
print(paste("Special Attack", data[248,"sp_attack"], "is in the", round(sp_attack_percentile(data[248,"sp_attack"]) * 100, 2), "th percentile"))
print(paste("Defense", data[248,"defense"], "is in the", round(defense_percentile(data[248,"defense"]) * 100, 2), "th percentile"))
print(paste("Special Defense", data[248,"sp_defense"], "is in the", round(sp_defense_percentile(data[248,"sp_defense"]) * 100, 2), "th percentile"))
print(paste("Speed", data[248,"speed"], "is in the", round(speed_percentile(data[248,"speed"]) * 100, 2), "th percentile"))
```

he has really good stats, particularly really high base attack and special attack, 
speed is not great, but has a lot of weaknesses.
however, many pokemon can learn abilities outside their two types, and Tyrannitar is no exception.
He can learn by TM flying, steel, fighting, fire, ice, type moves to counter its weaknesses.

stats distributions

```{r}
par(mfrow = c(1, 1))  # layout

stats <- c("hp", "attack", "defense", "sp_attack", "sp_defense", "speed")

overall_min <- min(c(data$speed, data$hp, data$attack, data$defense, data$sp_attack, data$sp_defense), na.rm = TRUE)
overall_max <- max(c(data$speed, data$hp, data$attack, data$defense, data$sp_attack, data$sp_defense), na.rm = TRUE)


plot(1, type = "n", xlab = "stats", ylab = "Frequency Density", xlim = c(overall_min, overall_max+10), ylim=c(0,0.019))

# plot each 
for (i in seq_along(stats)) {
  stat <- stats[i]
  density_values <- density(data[, stat], na.rm = TRUE)
  if (i == 2){
    lines(density_values$x, density_values$y, col = "goldenrod1")
  } else if (i == 4) {
    lines(density_values$x, density_values$y, col = "cyan3")
  } else {
    lines(density_values$x, density_values$y, col = rainbow(length(stats))[i])
  }
}

# Add legend
title("Distributions of main stats")
adapted_cols <- c("red","goldenrod1","green","cyan3","blue","magenta")
legend("topright", legend = stats, col = adapted_cols, lty = 1, cex = 0.8, pch=20)

# add point at max 
for (i in seq_along(stats)) {
  stat <- stats[i]
  density_values <- density(data[, stat], na.rm = TRUE)
  
  # find peak values where density is maximum
  peak_x <- density_values$x[which.max(density_values$y)]
  peak_y <- max(density_values$y)
  
  # Plot point at peak value
  if (i == 2) {
    points(peak_x, max(density_values$y), pch = 20, col = "goldenrod1", cex = 1.5)
  } else if (i == 4){
    points(peak_x, max(density_values$y), pch = 20, col = "cyan3", cex = 1.5)
  } else {
    points(peak_x, max(density_values$y), pch = 20, col = rainbow(length(stats))[i], cex = 1.5)
  }
  
  # Adjust text position for yellow and light blue points
  text_x <- peak_x
  text_y <- peak_y
  if (i == 2) {
    text_x <- peak_x +5  # Adjust the x-coordinate for better visibility
    text_color <- "goldenrod1"  # Use black text for better contrast
    # Add a line pointing to the dot
  } else if (i == 4){
    text_x <- peak_x -2
    text_color <- "cyan3"
  } else {
    text_x <- peak_x
    text_color <- rainbow(length(stats))[i]
  }
  
  #label
  text(text_x, text_y, label = sprintf("%.2f", peak_x), pos = 3, col = text_color, cex = 0.8)
}

```

```{r}
avg_stats_by_type <- function(data, target_type){
  subset_data <- data[data$type1 == target_type | data$type2 == target_type, ]
  avg_stats <- colMeans(subset_data[,main_stats])
  return(avg_stats)
}
all_types <- unique(c(data$type1, data$type2))
all_types <- all_types[all_types != "none"]
type_stats <- list()
# Loop 
for (type_to_test in all_types) {
  avg_stats <- avg_stats_by_type(data, type_to_test)
  type_stats[[type_to_test]] <- c(Type = type_to_test, avg_stats)
}

result_df <- bind_rows(type_stats)
result_df[, main_stats] <- lapply(result_df[, main_stats], as.numeric)

print(result_df)
```

```{r}
percentile_functions <- list(
  hp = hp_percentile,
  attack = attack_percentile,
  sp_attack = sp_attack_percentile,
  defense = defense_percentile,
  sp_defense = sp_defense_percentile,
  speed = speed_percentile
)

for (stat in main_stats) {
  cat("\nStat:", stat, "\n")
  
  percentile_function <- percentile_functions[[stat]]
  # Iterate over types
  for (type in all_types) {
    values <- result_df[[stat]][result_df$Type == type]
    
    cat(paste(stat, values, "is in the", 
              round(percentile_function(values) * 100, 2), "th percentile for", type), "\n")
  }
}
```

```{r}
densities <- density(data$hp)
plot(densities$x, densities$y, type="l", lwd=2, main="Percentiles of HP by type",xlab="HP",ylab="frequency density")
segments(x0=82, x1=82, y0=0,y1=0.011, col="slateblue1",lwd=3)
segments(x0=56.5,x1=56.5,y0=0, y1=0.016, col="yellow4", lwd=3)
text(x=130, y=0.012, col="slateblue1", cex=0.6,labels="dragon type is in the 76.53th percentile")
text(x=16, y=0.017, col="yellow4", cex=0.6, labels="bug type is in the 32.96th percentile")
```

```{r}
densities <- density(data$attack)
plot(densities$x, densities$y, type="l", lwd=2, main="Percentiles of Attack by type",xlab="Attack",ylab="frequency density")
segments(x0=103.3, x1=103.3, y0=0,y1=0.0076, col="tomato3",lwd=3)
segments(x0=62.9,x1=62.9,y0=0, y1=0.0124, col="rosybrown1", lwd=3)
text(x=152, y=0.008, col="tomato3", cex=0.8,labels="fighting type is in the 79.65th percentile")
text(x=18, y=0.012, col="rosybrown1", cex=0.8, labels="fairy type is in the 35.21th percentile")
```

